---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Nusrat Noor"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse) 
library(agricolae)
library(lubridate)
library(here)
#loaded the necessary packages
here()
#checked the working directory

Lter <- read.csv(here("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)
Lter$sampledate <- as.Date(Lter$sampledate, format = "%m/%d/%y")
#imported the raw file and set date columns to date objects

#2
my_theme <- theme_light(base_size = 14) +
  theme(axis.title = element_text(color = "darkgreen"), axis.text = element_text(color = "darkgreen"), legend.position = "right", legend.background = element_rect(fill = "darkgreen"), panel.background = element_rect(fill = "lightgrey"), panel.grid = element_line(color = "black"))
theme_set(my_theme)
#created theme and set it as default  

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature in July does not change with depth across all lakes.
Ha: Mean lake temperature in July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
Lter_wr <- Lter %>%
  filter(month(sampledate) == 07) %>% #filtered out dates in July
  select(lakename:daynum, depth, temperature_C) %>% #selected out just the desired columns
  na.omit(Lter) #got rid of any NAs

#5
Lter_tbd <- ggplot(Lter_wr, aes(x = depth, y = temperature_C)) +
  geom_point() +
  ylim(0, 35) +
  geom_smooth(method = "lm", color = "darkgreen") +
  labs(x = "Depth in Meters", y = "Temperature in Celsius")
print(Lter_tbd)
#made a scatterplot of temperature by depth and added a linear model line, added ylimits and changed the labels to make it look nicer
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that temperature decreases as depth increases. The trend does not seem linear because the temperature is distributed fairly wide in lower depths and then plateau's out after 10 meters. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
Lter_reg <- lm(data = Lter_wr, temperature_C ~ depth) #performed a linear regression
summary(Lter_reg) #checked the summary
par(mfrow = c(2,2), mar = c(4,4,4,4)) #set plot view to show all 4 plots at the same time
plot(Lter_reg) #plotted the regression
par(mfrow = c(1,1)) #reset the plot view
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The r-squared value is 0.7387 meaning that there is a strong correlation and that roughly 73.9% of the variability in temperature is explained by changes in depth. The degrees of freedom is 9726. The p-value is 2.2e-16 meaning it is statistically significant and we can reject the null hypothesis. For every 1 meter of depth, the temperature is expected to change by -1.946. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
TAIC <- lm(data = Lter_wr, temperature_C ~ year4 + daynum + depth) #performed linear regression 
step(TAIC) #conducted stepwise regression

#10
Treg <- lm(data = Lter_wr, temperature_C ~ year4 + daynum + depth) #ran multiple regression for the recommended set of variables
summary(Treg)
par(mfrow = c(2,2), mar = c(4,4,4,4))
plot(Treg)
par(mfrow = c(1,1))
#plotted the regression
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggested the use of the variables, year4, daynum, and depth. The r-squared increased to 0.7411 meaning that 74% of the observed variance is explained by the variables. This improved from the previous model but not by much. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Lter_anova <- aov(data = Lter_wr, temperature_C ~ lakename)
summary(Lter_anova)
Lter_anova2 <- lm(data = Lter_wr, temperature_C ~ lakename)
summary(Lter_anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The p-value is <2.2e-16 meaning there is a significant difference in mean temperature among the lakes and we would reject the null hypothesis. The different lakes on average have different temperatures in the month of July. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(Lter_wr, aes(x=depth, y=temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0,35) +
  labs(x = "Depth in Meters", y = "Temperature in Celcius", color = "Lakename") +
  theme(legend.text = element_text(color = "white"), legend.title = element_text(color = "white"))
  

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lter_anova) #computed Tukey HSD

Lter_anova_group <- HSD.test(Lter_anova, "lakename", group = TRUE) 
#extracted groupings for pairwise relationships
Lter_anova_group
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake was the only lake that had statistically same means as Peter Lake. Central Long Lake's mean was statistically distict from all the other lakes. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We would do a two sample t-test.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
Lter_wr2 <- filter(Lter_wr, lakename == "Crampton Lake" | lakename == "Ward Lake")
#filtered out only records for crampton lake and ward lake. 
Lter_2sample <- t.test(Lter_wr2$temperature_C ~ Lter_wr2$lakename) #conducted two sample t-test
Lter_2sample

```

>Answer: The p-value is 0.2649 meaning we would not reject the null hypothesis and the July temperatures are the same. This does match the results from question 16. 
