---
title: "Assignment 5: Data Visualization"
author: "Nusrat Noor"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

here()

Lake <- read.csv(here("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

Litter <- read.csv(here("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)

#2 
str(Lake)
Lake$sampledate <- ymd(Lake$sampledate)

str(Litter)
Litter$collectDate <- ymd(Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

my_theme <- theme_light(base_size = 14) + 
  theme(axis.title = element_text("black"), 
        plot.background = element_rect("gray"), 
        axis.text = element_text("black"))

theme_set(my_theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
LakePlot <- ggplot(Lake,aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  xlim(0, 50) +
  ylim(0, 150) +
  labs(x = "Phosphate (ug)", y = "Total Phosphorus (ug)", title = "Total Phosphorus by Phosphate", color = "Lake Name")
print(LakePlot)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
#why do we sometimes put aes in ggplot or in geom depending on the plot type?

temp_plot <- ggplot(Lake, aes(x = month.abb[month], y = temperature_C, color = lakename)) +
  geom_boxplot() +
  scale_x_discrete(limits = month.abb) +
  labs(x = "Month", y = "Temperature(C)", title = "Temperature by Month", color = "Lake Name")
print(temp_plot)

TP_plot <- ggplot(Lake, aes(x = month.abb[month], y = tp_ug, color = lakename)) +
  geom_boxplot() +
  scale_x_discrete(limits = month.abb) +
  labs(x = "Month", y = "Total Phosphorus(ug)", title = "Total Phosphorus by Month", color = "Lake Name")
print(TP_plot)

TN_plot <- ggplot(Lake, aes(x = month.abb[month], y = tn_ug, color = lakename)) +
  geom_boxplot() +
  scale_x_discrete(limits = month.abb) +
  labs(x = "Month", y = "Total Nitrogen(ug)", title = "Total Nitrogen by Month", color = "Lake Name")
print(TN_plot)


legendtemp <- get_legend(temp_plot)
legendTP <- get_legend(TP_plot)

plot_all <- plot_grid(temp_plot + theme(legend.position = "none"), 
                      TP_plot + theme(legend.position = "none"), 
                      TN_plot + theme(legend.position = "bottom"), 
                      nrow = 3, rel_heights = c(.7, .7, 1)) 
print(plot_all)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 
I notice that warmer months have higher levels of total phosphorus and total nitrogen and also have a wider spread. The average total nitrogen and phosphorus levels in Peter lake is higher and overall, has more spread. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
Needles1 <- ggplot(subset(Litter, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  labs(x = "Date", y = "Dry Mass", title = "Dry Mass of Needles by Date", color = "NLCD Class")
print(Needles1)

#7
Needles2 <- ggplot(subset(Litter, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass), ncol = 3) +
  labs(x = "Date", y = "Dry Mass", title = "Dry Mass of Needles by Date")
print(Needles2)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
Plot 7 is  more effective because it separates the data much more clearly so that the points not overlapping  as much as they are in plot 6. 