---
title: "Assignment 10: Data Scraping"
author: "Nusrat Noor"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(lubridate)
library(here)
library(rvest)

here()

mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
website <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')

website
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water_sys <- website %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()
water_sys

PWSID <- website %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
PWSID

Ownership <- website %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
Ownership

MGD_monthly <- website %>%
  html_nodes("th~ td+ td") %>%
  html_text()
MGD_monthly

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
MGD_df <- data.frame("Month" = rep(1:12),
                     "Year" = rep(2022,12),
                     "Max_withdrawls" = as.numeric(MGD_monthly))

MGD_df <- MGD_df %>%
  mutate(Water_system = !!water_sys, 
         PWSID = !!PWSID,
         Ownership = !!Ownership,
         Date = my(paste0(Month,"-",Year)),
         Month = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")) #%>%
  #arrange(desc(Date))
view(MGD_df)
#5 

ggplot(MGD_df,aes(x=Date, y = Max_withdrawls)) +
  geom_line() +
  labs(y="Maximum Day Use", title = "2022 Maximum Daily Use per Month", subtitle = "Nusrat Noor")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
#the_pwsid <- '03-32-010'
#the_year <- 2015
#scrape_url <- paste0(base_url, 'pwsid=', the_pwsid, '&year=', the_year)

scrape.it <-function(the_pwsid, the_year){
  the_website <- read_html(paste0(base_url, 'pwsid=', the_pwsid, '&year=', the_year))
  
  water_sys_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  PWSID_tag <- 'td tr:nth-child(1) td:nth-child(5)'
  ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  mgd_monthly_tag <- 'th~ td+ td'
  
  water_sys <- the_website %>%
  html_nodes(water_sys_tag) %>%
  html_text()
  PWSID <- the_website %>%
  html_nodes(PWSID_tag) %>%
  html_text()
  Ownership <- the_website %>%
  html_nodes(ownership_tag) %>%
  html_text()
  MGD_monthly <- the_website %>%
  html_nodes(mgd_monthly_tag) %>%
  html_text()
  Durham_df <- data.frame("Month" = rep(1:12), 
                       "Year" = rep(the_year,length(months)), 
                       "Max_withdrawls" = as.numeric(MGD_monthly)) %>%
    mutate(Water_system = !!water_sys, 
         PWSID = !!PWSID,
         Ownership = !!Ownership,
         Date = my(paste(Month,"-",Year)),
         Month = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"))
  return(Durham_df)
} 

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
the_df <- scrape.it('03-32-010', 2015)
view(the_df)
ggplot(the_df, aes(x=Month, y=Max_withdrawls)) +
  geom_line() +
  labs(y="Maximum Day Use (mgd)", title = "2015 Maximum Daily Use per Month", subtitle = "Durham, NC")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 

Ash_df <- scrape.it('01-11-010', 2015)
view(Ash_df)
ggplot(Ash_df, aes(x=Date, y=Max_withdrawls)) +
  geom_line() +
  labs(y="Maximum Day Use (mgd)", title = "2015 Maximum Daily Use per Month", subtitle = "Ashville")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 


```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
>
